---
title: "The SARA clinical trial"
author: "Albert Cobos"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.align = 'center')

# reading data
library(dplyr)
d <- readxl::read_excel("SARA_with_outcomes.xlsx") %>% 
  mutate(group = factor(group, levels = 1:2, labels = c("CA", "ADT")),
         sex = factor(sex, levels = 1:2, labels = c("male", "female")),
         PEP = factor(PEP, levels = 0:1, labels = c("AF-free", "AF")),
         SEP = factor(SEP, levels = 0:1, labels = c("AF-free", "AF")),
         NYHA = as.factor(NYHA),
         # converting character to Date
         rand_dt = as.Date(rand_dt),
         birth_dt = as.Date(birth_dt),
         # computing age at randomization
         age = floor(as.numeric(rand_dt - birth_dt)/365.25),
         bmi = weight / (height/100)^2)

# loading packages
library(mosaic)
library(ggformula)
library(patchwork)
library(compareGroups)

```

## Objectives and trial design

### Objective

To compare the effectiveness of CA vs. ADT in treating persistent AF.

### Trial design

- Multicenter, open, parallel-groups, randomized 

- Randomization to either CA or ADT, with 2:1 allocation ratio.

- 12-month follow-up. 

- All centres experienced in AF ablation procedures.

## Outcomes

### Primary outcome

- AF/flutter lasting more than 24 h or requiring cardioversion, during follow-up. 

### Secondary outcomes

- AF/flutter lasting at least 30 s, during follow-up. 

- Number of cardioversions during follow-up. 

- AF-QoL questionnaire (total score) at baseline, 6 and 12 months. 

## Results - Baseline characteristics {.smaller}


```{r pats_per_group}

# petients per group
ng <- tally(~group, data = d)

```

`r nrow(d)` patients randomized (`r paste(names(ng), ng, collapse = ", ")`), from `r format(min(d$rand_dt), "%B %Y")` to `r format(max(d$rand_dt), "%B %Y")`.

```{r baseline}

# labels for dataset variables
Hmisc::label(d$age) <- "Age at randomization (y)"
Hmisc::label(d$sex) <- "Sex"
Hmisc::label(d$bmi) <- "BMI (kg/m^2)"
Hmisc::label(d$SBP) <- "SBP (mmHg)"
Hmisc::label(d$DBP) <- "DBP (mmHg)"
Hmisc::label(d$HR) <- "Heart rate (bpm)"
Hmisc::label(d$QoL_bas) <- "AF-QoL score"

descrTable(group ~ age + sex + bmi + NYHA + SBP + DBP + HR + QoL_bas, data = d,
           show.p.overall = FALSE) %>% 
  export2md(caption = "")

```


## Results - Outcomes {.smaller}

```{r outcomes}

d <- d %>% 
  # recode cardioversions and compute AF-QoL change at 6 and 12 months
  mutate(Cardioversions = factor(ifelse(d$Cardioversions > 1, 
                                        "2++", 
                                        as.character(d$Cardioversions))),
         qol_change_6m = QoL_6m - QoL_bas,
         qol_change_12m = QoL_12m - QoL_bas)

Hmisc::label(d$PEP) <- "Free of AF or flutter (during >24 h)"
Hmisc::label(d$SEP) <- "Free of AF or flutter (during >30 s)"
Hmisc::label(d$Cardioversions) <- "Number of cardioversions"
Hmisc::label(d$qol_change_6m) <- "AF-QoL, 6 months change"
Hmisc::label(d$qol_change_12m) <- "AF-QoL, 12 months change"

descrTable(group ~ PEP + SEP + Cardioversions + qol_change_6m + qol_change_12m, data = d,
           show.p.overall = TRUE,
           hide.no = "AF") %>% 
  export2md(caption = "",
            header.labels = c('p.overall' = "P-value"))

```


## Results - AF-QoL {.smaller}

```{r qol_change_boxplots, fig.height=2, fig.cap="Changes from baseline in AF-QoL, at 6 and 12 months"}

p1 <- gf_boxplot(qol_change_6m ~ group, data = d,
                 xlab = "",
                 ylab = "AF-QoL, 6 months change") %>% 
  gf_refine(coord_flip()) %>% 
  gf_theme(theme_light())
  

p2 <- gf_boxplot(qol_change_12m ~ group, data = d,
                 xlab = "",
                 ylab = "AF-QoL, 12 months change") %>% 
  gf_refine(coord_flip()) %>% 
  gf_theme(theme_light())

p1 + p2


```

```{r qol_mean_diff}

# extracting results from the Welch t-tests for change at 6 months
r6 <- t.test(qol_change_6m ~ group, data = d, R = 5000)
r6_means <- r6$estimate
r6_mean_diff <- round(-diff(r6_means), 1)
names(r6_mean_diff) <- "difference (ADT-CA)"
r6_mean_diff_ci <- reporttools::displayCI(r6$conf.int, digit=1)

# extracting results from the Welch t-tests for change at 12 months
r12 <- t.test(qol_change_12m ~ group, data = d, R = 5000)
r12_means <- r12$estimate
r12_mean_diff <- round(-diff(r12_means), 1)
names(r12_mean_diff) <- "difference (ADT-CA)"
r12_mean_diff_ci <- reporttools::displayCI(r12$conf.int, digit=1)

```

\


- Change at 6 months: CA = `r r6_means["mean in group CA"] %>% round(1)`, ADT = `r r6_means["mean in group ADT"] %>% round(1)`, difference [95% CI]: `r r6_mean_diff` `r r6_mean_diff_ci`. 

- Change at 12 months: CA = `r r12_means["mean in group CA"] %>% round(1)`, ADT = `r r12_means["mean in group ADT"] %>% round(1)`, difference [95% CI]:`r r12_mean_diff` `r r12_mean_diff_ci`.

## Conclusions

- Catheter ablation was superior to medical therapy for the maintenance of sinus rhythm in patients with persistent AF at 12-month follow-up.
