---
title: "Birth weight"
author: "Albert Cobos"
date: "04/08/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.align = 'center')
                      
library(dplyr)

d <- MASS::birthwt %>%
  mutate(low = factor(low, 
                     levels = 1:0, 
                     labels = c("low", "normal")),
         race = factor(race, 
                       levels = 1:3,
                       labels = c("white", "black", "other")),
         smoke = factor(smoke, 
                        levels = 1:0, 
                        labels = c("smoker", "non-smoker")),
         ui = factor(ui, 
                      levels = 1:0, 
                      labels = c("UI", "no-UI")),
         ht = factor(ht, 
                     levels = 1:0, 
                     labels = c("AHT", "no-AHT"))) %>% 
  rename(bw = low)
```

## Introduction

We analysed the birth weight of `r nrow(d)` newborns recorded in the Baystate Medical Center, Springfield, Mass during 1986, and possible assciations with demographic characteristics of mothers. 
 
## Methods

Data are summarised as mean (SD) and n (%) as appropriate. For comparisons between groups the t-test and chisquare test with Yate'scontinuity correction were used, for quantitative and categorical variables respectively. The analysis was carried out with the R language (version 4.0.4). 

## Results

Birth weights ranged from `r min(d$bwt)` to `r max(d$bwt)` grams, with mean (SD) of `r round(mean(d$bwt),1)` (`r round(sd(d$bwt),1)`) grams. The histogram birth weights, with overlayed density, shows a bell-shaped and quite symmetrical distribution (figure \ref{fig:btw_density}).

```{r btw_density, fig.cap='Distribution of birth weights', fig.width=4, fig.height = 3}

library(ggformula)

gf_dens(~ bwt, data = d, 
        color = "blue", size = 1,
        xlab = "Birth weight (grams)") %>% 
  gf_histogram(binwidth=500, color = "black", fill = "white") %>% 
  gf_theme(theme_light())

```

The boxpots of birth weights in smoking and non-smoking mothers are dispalyed in figure \ref{fig:bwt_smoke_boxplot}. A shift to lower values is apparent in smoking mothers. 


```{r bwt_smoke_boxplot, fig.cap='Boxplots of birth weight by smoking status', fig.height=1.5, fig.width=4 }

gf_boxplot(bwt ~ smoke, data = d,
           ylab = "Birth weight (grams)",         # defines a labe for the x-axis
           xlab = "") %>%                         # defines a void label for the y-axis
  gf_refine(coord_flip()) %>%                     # flips the axes
  gf_theme(theme_light())                         # uses the light 'theme' (or style)

```

Table \ref{tab:smoke_comparisons} shown the results of comparing several dataset variables in smoker and non-smoker mothers. As well as descriptive summaries, the p-value of the  relevant test (t-test for quantitative variables, and chi-square for categorical variables) is shown as `p.overall`.


```{r smoke_comparisons, results='markup'}

library(compareGroups)

Hmisc::label(d$bwt) <- "Birth weight (grams)"
Hmisc::label(d$bw) <- "Birth weight, categorized"
Hmisc::label(d$race) <- "Race"
Hmisc::label(d$ht) <- "Arterial hypertension"
Hmisc::label(d$ui) <- "Uterine irritability"

descrTable(smoke ~ bwt + bw + race + ht + ui, data = d,
           hide.no = c("no-AHT", "no-UI"))  %>%   
  export2md()                                                 

```

